---
title: "HW1"
author: "Ruby Krasnow"
date: "2024-08-22"
link-external-newwindow: true 
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: load-packages
#| message: FALSE
#| echo: FALSE
#| warning: FALSE
#| output: FALSE

# List of packages required:
packages <- c("tidyverse", "PNWColors", "janitor")

# Load packages into session
lapply(packages, require, character.only = TRUE)
rm(packages)

# Ensure functions with duplicate names are from the correct package
select <- dplyr::select
map <- purrr::map
summarize <- dplyr::summarize
clean_names <- janitor::clean_names
margin <- ggplot2::margin

set.seed(123) #Set seed for pseudo-random number generator, for reproducibility
```

# 1) Who am I & what do I study?

I am an aspiring marine ecologist with a strong interest in sustainable fisheries and aquaculture. I am especially passionate about using mathematical and statistical models to understand aquatic ecosystems, support sustainable fisheries management, and advance the kelp and shellfish aquaculture industries. I am currently a senior at Clark University, where I am majoring in Biology with a minor in Mathematics. After graduating with my B.A. in December 2024, I will start my PhD in Marine Biology at the University of Maine. I am one of the team captains for the [Clark cross-country team](https://clarkathletics.com/sports/womens-cross-country) and am excited to continue my athletic as well as academic career by competing on the UMaine track and cross-country teams as a graduate student.

A major focus of my research is enhancing the accuracy of kelp (Saccharina latissima) growth models by incorporating factors such as blade erosion, biofouling, genetic variation, and phenotypic plasticity. Additionally, I am collaborating with the NOAA Northeast Fisheries Science Center to model spatial variation in crustacean size-at-maturity, supporting the sustainable management of the emerging Jonah crab fishery in New England.

To learn more about me, you can visit my personal website: <https://rmk118.github.io/>

### Keywords

-   quantitative marine ecology
-   fisheries & aquaculture
    -   kelp aquaculture
    -   modeling crustacean size-at-maturity
    -   population dynamics & stock assessment
-   synthesis research/meta-analyses
-   open & reproducible data science

# 2) What do I want from this course?

1.  Develop a more solid theoretical understanding of the concepts underlying the modeling techniques I frequently use in my research
2.  Learn new techniques for model selection and inference (e.g., practice cross-validation)
3.  Become more familiar with distributions/models that can address common issues in ecological analyses, such as zero-inflated and hurdle models

# 3) Plot air quality data

```{r}
#| label: setup-data
#| echo: false

air <- airquality %>% 
  clean_names() %>% 
  select(wind, temp)
```

```{r}
#| label: fig-windtemp
#| fig-cap: "Avg. wind speed (mph) and max. daily temperature (°F) at LaGuardia Airport, NY, from May-Sept. 1973"
#| fig-width: 5
#| fig-height: 4
#| fig-align: "center"

ggplot(air)+
  geom_point(aes(x=wind, y=temp))+
  theme_light()+ #define custom theme for ggplots
  theme(
    axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0)),
    axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0)),
    text=element_text(size=13))+labs(x="Wind speed (mph)", y="Temperature (°F)")+
  xlim(0,NA)
```

# 4) Say it with an equation

$$ Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \varepsilon $$ {#eq-linmod}
$$ \varepsilon \sim N(0, \sigma^2)$$

In @eq-linmod, the two different predictor variables are represented by $X_1$ and $X_2$.