---
title: "HW2: Linear Models"
author: "Ruby Krasnow"
link-external-newwindow: true 
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: load-packages
#| message: FALSE
#| echo: FALSE
#| warning: FALSE
#| output: FALSE

# List of packages required:
packages <- c("tidyverse", "PNWColors", "janitor", "faraway", "broom", "DHARMa", "emmeans", "performance")

# Load packages into session
lapply(packages, require, character.only = TRUE)
rm(packages)

# Ensure functions with duplicate names are from the correct package
select <- dplyr::select
map <- purrr::map
summarize <- dplyr::summarize
clean_names <- janitor::clean_names
margin <- ggplot2::margin

set.seed(123) #Set seed for pseudo-random number generator, for reproducibility

mytheme <- theme_light()+ #define custom theme for ggplots
  theme(axis.title.y = element_text(margin = margin(t = 0, r = 10, l = 0)),
        axis.title.x = element_text(margin = margin(t = 10, l = 0)),
        text=element_text(size=15))

```

[*Link to original assignment PDF*](https://qerm514.github.io/website/homework/week_02/hw_02_fitting_lm.pdf)

# Background

The goal of this assignment is to familiarize yourself with fitting linear models in **R**. We will be working some data from nearby Lake Washington that is part of a long-term monitoring program begun in the 1960s by the late, and rather famous, [Dr. W.T. Edmondson](http://www.nasonline.org/publications/biographical-memoirs/memoir-pdfs/edmondson-wt.pdf) and since continued by [Dr. Daniel Schindler](https://fish.uw.edu/faculty/daniel-schindler/). The accompanying data file `L_Washington_plankton.csv` contains information on the following four variables:

* `Daphnia`: index of the density of the cladoceran *Daphnia* (unitless)

* `Greens`: index of the density of green algae (unitless)

* `Cyclops`: index of the density of the copepod *Cyclops* (unitless)

* `Temp`: water temperature (°C)

*Daphnia* are an effective grazer on phytoplankton and green algae make up a large proportion of their diet. *Cyclops* are an inferior grazer compared to *Daphnia*, but a competitor nonetheless. *Daphnia* growth rates are also affected by water temperature.


# Question 1

a) Write out the equation for a linear regression model that expresses *Daphnia* abundance as a function of its preferred prey, green algae, and describe the terms in your model.

$$y_i =\beta_0 + \beta_1 x_i+e_i $$
Each observation of *Daphnia* abundance $(y_i)$ is a function of an intercept $(\beta_0)$ and the effect $(\beta_1)$ of green algae density $(x_i)$. The model residuals are assumed to be independent and normally distributed with mean 0 and variance $\sigma^2$, such that $e_i \sim \mathrm{N}(0, \sigma^2)$.

b) Produce a scatterplot that shows the relationship between `Daphnia` and `Greens`. Make sure to label your plot accordingly and give it an informative caption. Describe the relationship between `Daphnia` and `Greens`. Does a linear model seem reasonable here?

```{r}
#| label: fig-daph1
#| fig-cap: Relationship between the density of *Daphnia* and its preferred prey, green algae
#| echo: false

daph <- read.csv("../data/L_Washington_plankton.csv")

ggplot(data=daph)+
  geom_point(aes(x=Greens, y=Daphnia))+mytheme+
  labs(x="Density of green algae", y=expression(paste("Density of ", italic("Daphnia"))))

```

There seems to be a weak positive relationship between green algae and *Daphnia* density that could be approximated with a linear model. 

c) Produce the step-by-step **R** code required to fit your model **via linear algebra** to generate estimates the model parameters and the data. Be sure to show the construction of the design matrix $(\mathbf{X})$, the calculation of the parameter estimates $(\hat{\beta_i})$, the calculation of the hat matrix $(\mathbf{H})$, and the calculation of the model predictions $(\hat{y_i})$.

First, we construct the design matrix:
```{r}
#| label: design-matrix

nn <- nrow(daph)
yy <- matrix(data=daph$Daphnia, nrow = nn, ncol = 1)
head(yy)

intercept <- rep(1, nn)
greens <- daph$Greens
XX <- cbind(intercept, greens)
head(XX)
```

Using the design matrix, we can find the parameter estimates $(\hat{\beta_i})$ using the following formula:
$$
\hat{\boldsymbol{\beta}} = (\mathbf{X}^{\top} \mathbf{X})^{-1} \mathbf{X}^{\top} \mathbf{y}.
$$
```{r}
beta_hat <- solve(t(XX) %*% XX) %*% t(XX) %*% yy # or solve(crossprod(XX,XX),crossprod(XX,yy))
beta_hat

#check answer
daph_mod <- lm(data=daph, Daphnia~Greens)
tidy(daph_mod) %>% select(c(term, estimate))
```

Now we find the hat matrix $(\mathbf{H})$:
```{r}
## hat matrix
HH <- XX %*% solve(t(XX) %*% XX) %*% t(XX)
dim(HH)

# Check answer
hat_auto <- optR::hatMatrix(XX)
all.equal(HH, hat_auto)
```

Finally, we find the model predictions $(\hat{y_i})$:
```{r}
y_hat <- HH %*% yy
#or
y_hat <- XX %*% beta_hat

# alternative built-in methods
predict(daph_mod)
head(broom::augment(daph_mod)) # see fitted column
```


d) Calculate and report your estimate of the residual variance $(\sigma^2)$.

We can use the following formula:
$$\sigma^2 = \frac{\mathrm{RSS}}{n-p}, \quad\text{where}$$
$$\mathrm{RSS}= \mathbf{e^\top}\mathbf{e}=\left(\mathbf{y}-X\hat{\beta}\right)^\top\left(\mathbf{y}-X\hat{\beta}\right)$$
```{r}
rss <- as.numeric(crossprod(yy-XX %*% beta_hat) )
rss

# check answer
all.equal(rss, deviance(daph_mod))

sigma2 <- rss/(nn-2)
sigma2

# check answer
all.equal(sigma2, sigma(daph_mod)^2)
```

e) Give a prediction of what you might expect the specific abundance of *Daphnia* to be on the next sampling occasion if the abundance of green algae is 1.5 units. Also provide an estimate of the interval around your estimate that conveys 95% confidence in your prediction. Again, do so via direct calculations rather than relying on **R**'s built-in functions.

First, we use the $\beta_0$ and $\beta_1$ values we found above to estimate $y$ when $x=1.5$.
```{r}
# y = b_0 + b_1 * x
y_star <- unname(beta_hat[1,1])+unname(beta_hat[2,1])*1.5
y_star
#OR

#vector of new data
X_star <- c(intercept = 1, greens = 1.5)
sum(X_star * beta_hat)

```

Now we need to find the standard error of our estimate
```{r}
## inside sqrt
inner_X <- t(X_star) %*% solve(t(XX) %*% XX) %*% X_star

## critical t-value
t_crit <- qt(0.975, df = nn-2)

## estimated SD
sigma <- sqrt(sigma2)

## 95% CI
y_star + c(-1,1) * t_crit * sigma * c(sqrt(inner_X))

# for prediction interval, 
inner_X_pred <- 1+ t(X_star) %*% solve(t(XX) %*% XX) %*% X_star
y_star + c(-1,1) * t_crit * sigma * c(sqrt(inner_X_pred))
```

Now let's check our answer using the built-in functions in **R**
```{r}
predict(daph_mod, new = data.frame(Greens=1.5),
        level = 0.95, interval = "confidence")

predict(daph_mod, new = data.frame(Greens=1.5),
        level = 0.95, interval = "pred")
```


# Question 2

a) Expand upon your model from Question 1 to include the additional effects of *Cyclops* and water temperature on *Daphnia*. Write out your equation and describe the terms in the model.

$$y_i =\beta_0 + \beta_1 x_{1,i}+\beta_2 x_{2,i}+\beta_3 x_{3,i}+e_i $$
Each observation of *Daphnia* abundance $(y_i)$ is a function of an intercept $(\beta_0)$, the effect $(\beta_1)$ of green algae density $(x_{1,i})$, the effect $(\beta_2)$ of *Cyclops* density $(x_{2,i})$, and the effect $(\beta_3)$ of water temperature $(x_{3,i})$. The model residuals are again assumed to be independent and normally distributed with mean 0 and variance $\sigma^2$, such that $e_i \sim \mathrm{N}(0, \sigma^2)$.

b) Using **R**'s built-in functions, fit the model from (a) and show the resulting table of results. For each of the $p$-values shown in the table, describe the null hypothesis being tested.

```{r}
daphmod_full <- lm(Daphnia ~ Greens + Cyclops + Temp, data=daph)
sumary(daphmod_full)
```

The first $p$-value corresponds to a t-test that the intercept is 0, $H_0: \beta_0=0$. The remaining three $p$-values shown in the table correspond to null hypothesis tests that the given predictor can be dropped from the model.

For example, can `Cyclops` be dropped from this model? We fit a reduced model and compare to it to the full model via an $F$-test with $H_0: \beta_\text{cyclops} = 0$

$$
\begin{aligned}
\Theta: \text{Daphnia}_i &= \beta_0 + \beta_1 \text{Greens}_i + \beta_2 \text{Cyclops}_i + \beta_3 \text{Temp}_i + e_i \\
~ \\ 
\theta: \text{Daphnia}_i &= \beta_0 + \beta_1 \text{Greens}_i + \beta_2 \text{Temp}_i + e_i
\end{aligned}
$$

```{r}
## reduced model without `Cyclops`
reduced_mod <- lm(Daphnia ~ Greens + Temp, data=daph)
## use `anova('reduced', 'full')` to get the F-test results
anova(reduced_mod, daphmod_full)
```

We see that the resulting p-value is the same as the one displayed for the Cyclops estimate in the summary table for the full model.

c) Test the hypothesis that $\beta_{Greens} = \beta_{Cyclops} = \beta_{Temp} = 0$. What is the $F$-statistic, the associated $df$, and the $p$-value? What can you conclude from this test?


We write the null hypothesis as
$$H_0: \beta_1 = \beta_2 = \beta_3 = 0$$
which corresponds to a model where we simply estimate the data based on their mean.

\begin{align*}
\Theta: \mathbf{y} &= \mathbf{X} \boldsymbol{\beta} + \mathbf{e} \\
\theta: \mathbf{y} &= \boldsymbol{\mu} + \mathbf{e} \\
\end{align*}

We will base this test on an $F$-distribution, such that


\begin{align*}
SSE_{\Theta} &= \left( \mathbf{y} - \mathbf{X} \boldsymbol{\beta} \right)^{\top} \left( \mathbf{y} - \mathbf{X} \boldsymbol{\beta} \right) = \mathbf{e}^{\top} \mathbf{e} = SSE \\
SSE_{\theta} &= \left( \mathbf{y} - \bar{y} \right)^{\top} \left( \mathbf{y} - \bar{y} \right) =  SSTO \\
&\Downarrow \\
F &= \frac{ \left( SSTO - SSE \right)  / (k - 1) } { SSE  / (n - k)}
\end{align*}

#### $F$-test by hand


```{r F_test_by_hand}
## get matrix of predictors
XX <- daph %>% select(-Daphnia) %>% add_column(Intercept=1, .before = "Temp") %>% as.matrix()
## estimate beta
beta_hat <- solve(t(XX) %*% XX) %*% t(XX) %*% yy
## total sum of squares
SSE <- t(yy - XX %*% beta_hat) %*% (yy - XX %*% beta_hat)
## error sum of squares
SSTO <- t(yy - mean(yy)) %*% (yy - mean(yy))
## F statistic
(F_stat <- ((SSTO - SSE) / (4 - 1)) / (SSE / (nn - 4)))
## F test
pf(F_stat, 4-1, nn-4, lower.tail = F)
```

This $F$-statistic is quite large and the $p$-value is very small, so we would reject the null hypothesis that we would be justified in dropping the 3 predictors from this model in favor of a mean-only model.

Checking using the built-in summary function, we see that the F-statistic and p-value reported at the bottom of the output are the same as our manually calculated values.
```{r}
summary(daphmod_full)
```


d) It has come to your attention that someone has done lab experiments suggesting the effect of temperature on *Daphnia* abundance is 0.4 per degree Celsius after controlling for the effects of prey (green algae) and competitors (*Cyclops*). Create a null hypothesis test to evaluate the evidence for this finding from the data collected in the field. Specify $H_0$ and report the results of your test. What do you conclude?

$$
\text{Daphnia}_i = \beta_0 + \beta_1 \text{Greens}_i + \beta_2 \text{Cyclops}_i + 0.4* \text{Temp}_i + e_i
$$

We write the null hypothesis as
$$H_0: \beta_{\text{temp}} = 0.4$$
```{r}
## model with effect of `elevation` = 1
fixed_mod <- lm(Daphnia ~ Greens + Cyclops + offset(0.4 * Temp), data=daph)
## use `anova('comb', 'full')` to get the F-test results
anova(fixed_mod, daphmod_full)
```

We have found significant evidence to reject $H_0$, suggesting that it may be worthwhile to re-run the lab experiments demonstrating the effect of temperature on *Daphnia* abundance is 0.4/°C after controlling for green algae and *Cyclops* density and determine if the findings can be replicated.